---
title: "HOPR"
author: "GRACE JUAN"
format: html
editor: visual
---

# **Chapter 1** Getting started with R and RStudio

```{r}
install.packages('remotes', dependencies = TRUE)
```

```{r}
update.packages(ask = FALSE)
```

```{r}
plot(1:10)
```

```{r}
install.packages('BiocManager', dependencies = TRUE)
```

```{r}
BiocManager::install()
```

```{r}
BiocManager::install(c("GenomicRanges", "edgeR"))
```

```{r}
BiocManager::install(ask = FALSE)
```

```{r}
remotes::install_github('tidyverse/dplyr')
```

# **Chapter 2** Some R basics

```{r}
2 + 2
## [1] 4
```

```{r}
log(1)              # logarithm to base e
## [1] 0
log10(1)            # logarithm to base 10
## [1] 0
exp(1)              # natural antilog
## [1] 2.718282
sqrt(4)             # square root
## [1] 2
4^2                   # 4 to the power of 2
## [1] 16
pi                    # not a function but useful
## [1] 3.141593
```

```{r}
my_obj <- 48
```

```{r}
my_obj
## [1] 48
```

```{r}
my_obj2 <- 1024
```

```{r}
my_obj3 <- my_obj + my_obj2
my_obj3
```

```{r}
output_summary <- "my analysis"
output.summary <- "my analysis"
outputSummary <- "my analysis"
```

```{r}
my_vec <- c(2, 3, 1, 6, 4, 3, 3, 7)
```

```{r}
my_vec
```

```{r}
mean(my_vec)    # returns the mean of my_vec
## [1] 3.625
var(my_vec)     # returns the variance of my_vec
## [1] 3.982143
sd(my_vec)      # returns the standard deviation of my_vec
## [1] 1.995531
length(my_vec)  # returns the number of elements in my_vec
## [1] 8
```

```{r}
in_vec <- c(3, 1, 10, 7)
my_seq7 <- rep(in_vec, each = 3)    # repeats each element of 
                                  # the series 3 times
my_seq7
##  [1]  3  3  3  1  1  1 10 10 10  7  7  7
```

```{r}
my_seq7 <- rep(c(3, 1, 10, 7), each = 3) # repeats each 
                                         # element of the 
                                         # series 3 times
my_seq7
##  [1]  3  3  3  1  1  1 10 10 10  7  7  7
```

```{r}
my_seq6 <- rep(1:5, each = 3)   # repeats each element of the 
                              #series 3 times
my_seq6
##  [1] 1 1 1 2 2 2 3 3 3 4 4 4 5 5 5
```

```{r}
my_vec        # remind ourselves what my_vec looks like
## [1] 2 3 1 6 4 3 3 7
my_vec[3]     # extract the 3rd value
## [1] 1

# if you want to store this value in another object
val_3 <- my_vec[3]
val_3
## [1] 1
```

```{r}
my_vec[my_vec >= 4]        # values greater or equal to 4
## [1] 6 4 7
my_vec[my_vec < 4]         # values less than 4
## [1] 2 3 1 3 3
my_vec[my_vec <= 4]        # values less than or equal to 4
## [1] 2 3 1 4 3 3
my_vec[my_vec == 4]        # values equal to 4
## [1] 4
my_vec[my_vec != 4]        # values not equal to 4
## [1] 2 3 1 6 3 3 7
```

```{r}
# replace the 6th and 7th element with 100
my_vec[c(6, 7)] <- 100
my_vec
## [1]   2   3   1 500   4 100 100   7

# replace element that are less than or equal to 4 with 1000
my_vec[my_vec <= 4] <- 1000
my_vec
## [1] 1000 1000 1000  500 1000  100  100    7
```

```{r}
height <- c(180, 155, 160, 167, 181)
height
## [1] 180 155 160 167 181

p.names <- c("Joanna", "Charlotte", "Helen", "Karen", "Amy")
p.names
## [1] "Joanna"    "Charlotte" "Helen"     "Karen"     "Amy"
```

```{r}
# create a second vector
my_vec3 <- c(17, 15, 13, 19, 11, 0)

# add both vectors
my_vec2 + my_vec3
## [1] 20 20 20 20 20 20

# multiply both vectors
my_vec2 * my_vec3
## [1] 51 75 91 19 99  0
```

```{r}
mean_temp <- mean(temp, na.rm = TRUE)
mean_temp
## [1] 6.2875
```

```{r}
help.search("mean")
```

# **Chapter 3** Data in R

```{r}
num <- 2.2
class(num)
## [1] "numeric"

char <- "hello"
class(char)
## [1] "character"

logi <- TRUE
class(logi)
## [1] "logical"
```

```{r}
# coerce numeric to character
class(num)
## [1] "numeric"
num_char <-  as.character(num)
num_char
## [1] "2.2"
class(num_char)
## [1] "character"

# coerce character to numeric!
class(char)
## [1] "character"
char_num <- as.numeric(char)
## Warning: NAs introduced by coercion
```

```{r}
my_mat <- matrix(1:16, nrow = 4, byrow = TRUE)
my_mat
##      [,1] [,2] [,3] [,4]
## [1,]    1    2    3    4
## [2,]    5    6    7    8
## [3,]    9   10   11   12
## [4,]   13   14   15   16

my_array <- array(1:16, dim = c(2, 4, 2))
my_array
## , , 1
## 
##      [,1] [,2] [,3] [,4]
## [1,]    1    3    5    7
## [2,]    2    4    6    8
## 
## , , 2
## 
##      [,1] [,2] [,3] [,4]
## [1,]    9   11   13   15
## [2,]   10   12   14   16
```

```{r}
rownames(my_mat) <- c("A", "B", "C", "D")
colnames(my_mat) <- c("a", "b", "c", "d")
my_mat
##    a  b  c  d
## A  1  2  3  4
## B  5  6  7  8
## C  9 10 11 12
## D 13 14 15 16
```

```{r}
mat.1 <- matrix(c(2, 0, 1, 1), nrow = 2)    # notice that the matrix has been filled 
mat.1                                     # column-wise by default
##      [,1] [,2]
## [1,]    2    1
## [2,]    0    1

mat.2 <- matrix(c(1, 1, 0, 2), nrow = 2)
mat.2
##      [,1] [,2]
## [1,]    1    0
## [2,]    1    2

mat.1 + mat.2           # matrix addition
##      [,1] [,2]
## [1,]    3    1
## [2,]    1    3
mat.1 * mat.2           # element by element products
##      [,1] [,2]
## [1,]    2    0
## [2,]    0    2
mat.1 %*% mat.2         # matrix multiplication
##      [,1] [,2]
## [1,]    3    2
## [2,]    1    2
```

```{r}
list_2 <- list(colours = c("black", "yellow", "orange"), 
               evaluation = c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE), 
               time = matrix(1:6, nrow = 3))
list_2
## $colours
## [1] "black"  "yellow" "orange"
## 
## $evaluation
## [1]  TRUE  TRUE FALSE  TRUE FALSE FALSE
## 
## $time
##      [,1] [,2]
## [1,]    1    4
## [2,]    2    5
## [3,]    3    6
```

```{r}
p.height <- c(180, 155, 160, 167, 181)
p.weight <- c(65, 50, 52, 58, 70)
p.names <- c("Joanna", "Charlotte", "Helen", "Karen", "Amy")

dataf <- data.frame(height = p.height, weight = p.weight, names = p.names)
dataf
##   height weight     names
## 1    180     65    Joanna
## 2    155     50 Charlotte
## 3    160     52     Helen
## 4    167     58     Karen
## 5    181     70       Amy
```

```{r}
str(dataf)   
## 'data.frame':    5 obs. of  3 variables:
##  $ height: num  180 155 160 167 181
##  $ weight: num  65 50 52 58 70
##  $ names : chr  "Joanna" "Charlotte" "Helen" "Karen" ...
```

# **Chapter 4** Graphics with base R

Summarising your data, either numerically or graphically, is an important (if often overlooked) component of any data analysis. Fortunately, R has excellent graphics capabilities and can be used whether you want to produce plots for initial data exploration, model validation or highly complex publication quality figures. There are three main systems for producing graphics in R; base R graphics, lattice graphics and ggplot2. Each of these systems have their strengths and weaknesses and we often use them interchangeably. In this Chapter we\'ll focus mostly on base R graphics with a sprinkling of lattice graphics for added variety. In the next [Chapter](https://intro2r.com/graphics_r.html#graphics_r) we\'ll introduce you to the `ggplot2` package.

The base R graphics system is the original plotting system that\'s been around (and has evolved) since the first days of R. When creating plots with base R we tend to use high level functions (like the `plot()` function) to first create our plot and then use one or more low level functions (like `lines()` and `text()` etc) to add additional information to these plots. This can seem a little weird (and time consuming) when you first start creating fancy plots in R, but it does allow you to customise almost every aspect of your plot and build complexity up in layers. The flip side to this flexibility is that you\'ll often need to make many decisions about how you want your plot to look rather than rely on the software to make these decisions for you. Having said that, it\'s generally very quick and easy to generate simple exploratory plots with base R graphics.

The lattice system is implemented in the `lattice()` package that comes pre-installed with the standard installation of R. However, it won\'t be loaded by default so you\'ll first need to use `library(lattice)` to access all the plotting functions. Unlike base R graphics, lattice plots are mostly generated all in one go using a single function so there\'s no need to use high and low level plotting functions to customise the look of a plot. This can be a real advantage as things like margin sizes and plot spacing are adjusted automatically. Lattice plots also make a few more decisions for you about how the plots will look but this comes with a slight cost as customising lattice plots to get them to look exactly how you want can become quite involved. Where lattice plots really shine is plotting complex multi-dimensional data using panel plots (also called trellis plots). We\'ll see a couple of examples of these types of plots later in the Chapter.

There are many functions in R to produce plots ranging from the very basic to the highly complex. It\'s impossible to cover every aspect of producing graphics in R in this introductory book so we\'ll introduce you to most of the common methods of graphing data and describe how to customise your graphs later on in this [Chapter](https://intro2r.com/custom_plot.html#custom_plot).

All of the plots we\'ve created so far in this Chapter are more than suitable for exploring your data. If however, you\'d like to make them a little prettier (for your thesis, publication or even your own amusement) you\'ll need to invest some time learning how to customise your plots. The good news is that the base R graphics system allows you to change almost any aspect of your plot. There are however a couple of things to bear in mind. Firstly, although many of the approaches we introduce in this section will work with most base R plotting functions, there\'s no true consistency between functions. What works with the `plot()` function isn\'t guaranteed to necessarily work with the `boxplot()` function. This can be a little frustrating to begin with but gets easier the more experience you gain. If you crave a little more consistency take a look at [Chapter 5](https://intro2r.com/graphics_r.html#graphics_r) where we introduce the excellent `ggplot2` package. Secondly, when you start customising plots you\'re confronted with a huge number of options and arguments to try and remember. This isn\'t necessarily a bad thing as this is what makes base R graphics so flexible but it\'s a lot to take in. Often a quick Google or peek at the relevant help pages will jog your memory. Thirdly, learning how to customise plots in base R isn\'t just about what code you need to use, it\'s also about learning the process of building a plot. We often start with a basic layout of our plot and then add layers of complexity until we achieve the desired results. This requires a little experience (and trial and error), but again becomes easier with practice. Lastly, this section covers the basics of how to customise base R graphics and most (if not all) of these approaches will not work for plots created with the `lattice` graphics system.

There are a number of different methods for plotting multiple graphs within the same graphics device, some of which you\'ve already met such as `pairs()`, `coplot()`, `xyplot()` etc. However these functions rely on plotting multiple graphs in different panels within the same plot. If you want to plot separate plots within the same graphics device you\'ll need a different approach. One of the most common methods is to use the main graphical function `par()` to split the plotting device up into a number of defined sections using the `mfrow =` argument. With this method, you first need to specify the number of rows and columns of plots you would like and then run the code for each plot. For example, to plot two graphs side by side we would use `par(mfrow = c(1, 2))` to split the device into 1 row and two columns.

Creating plots in R is all well and good but what if you want to use these plots in your thesis, report or publication? One option is to click on the \'Export\' button in the \'Plots\' tab in RStudio as we described [previously](https://intro2r.com/baseplot_start.html#baseplot_start). You can also export your plots from R to an external file by writing some code in your R script. The advantage of this approach is that you have a little more control over the output format and it also allows you to generate (or update) plots automatically whenever you run your script. You can export your plots in many different formats but the most common are, pdf, png, jpeg and tiff.

By default, R (and therefore RStudio) will direct any plot you create to the plot window. To save your plot to an external file you first need to redirect your plot to a different graphics device. You do this by using one of the many graphics device functions to start a new graphic device. For example, to save a plot in pdf format we will use the `pdf()` function. The first argument in the `pdf()` function is the filepath and filename of the file we want to save (don\'t forget to include the .pdf extension). Once we\'ve used the `pdf()` function we can then write all of the code we used to create our plot including any graphical parameters such as setting the margins and splitting up the plotting device. Once the code has run we need to close the pdf plotting device using the `dev.off()` function.

# **Chapter 5** Graphics with ggplot
